---
title: "Bulk RNA Analysis"
subtitle: "Wrangle"
author: "Wes Horton"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
---

```{r loadLibs}
### Load Utils package
if (!require("wrh.rUtils")) remotes::install_github("weshorton/wrh.rUtils")

library(data.table)
library(wrh.rUtils)
```

# Load Data

Set the main directory as well as the sub-directory for the input files

```{r set directories}
baseDir_v <- "~/my_tool_repos/humanValidation/"
dataDir_v <- mkdir(baseDir_v, "inputs")
```

Specify file names for counts and meta

```{r file names}
countName_v <- "j15221_evanthia_breast_rnaseq_deidentified.tsv"
metaName_v <- "table_24patient_info.csv"
```

Read them in (data.table)

```{r read}
counts_dt <- fread(file.path(dataDir_v, countName_v))
meta_dt <- fread(file.path(dataDir_v, metaName_v))
```

# Harmonize Samples

Depending on the project, the metadata may describe more samples than are present in the count data.  

Process will depend on the specific project, but at the end of this, the metadata and the count data should have the same samples.

```{r harmonize}
# Sample names in the count data are a combination of the patient ID and Timepoint, separated by underscore
# Sample names in the metadata are simply the patient ID

# Get unique sample names from RNAseq data
countSamples_v <- unique(gsub("_.*$", "", grep("^R", colnames(counts_dt), value = T)))

# Check
inCountNotMeta_v <- setdiff(countSamples_v, meta_dt$ID) # Should be empty
inMetaNotCount_v <- setdiff(meta_dt$ID, countSamples_v)

# Update
cat(sprintf("%s samples are in the metadata sheet, but don't have RNAseq data:\n\t%s", length(inMetaNotCount_v), paste(inMetaNotCount_v, collapse = "\n\t")))

# Remove from meta
meta_dt <- meta_dt[!(ID %in% inMetaNotCount_v),]
```

# Wrangle Meta for DESeq

The metadata only has one row per sample while the count data has multiple columns per sample (multiple timepoints).  

DESeq functions require the metadata rows to exactly match the count columns, so the metadata will need to be "duplicated".  

Since each patient has a variable number of timepoints (i.e. not all patients have all timepoints), make them all and then subset.

```{r}
# Specify the timepoints for the project
timePoints_v <- c("Baseline", "Post Run-in", "Week 8")

# Specify the gene column name (this is the only column in the count table that isn't a measurement)
geneCol_v <- "gene"

# Copy the original ID (this will no longer be unique)
meta_dt$patientID <- meta_dt$ID

# Iterate over each timepoint, replacing the original ID with new ID_Timepoint
meta_lsdt <- list()
for (tp_v in timePoints_v) {
  tmpMeta_dt <- copy(meta_dt)
  tmpMeta_dt$Timepoint <- tp_v
  tmpMeta_dt$ID <- paste(tmpMeta_dt$ID, tp_v, sep = "_")
  meta_lsdt[[tp_v]] <- tmpMeta_dt
}

# Combine them all together
allMeta_dt <- do.call(rbind, meta_lsdt)

# Subset to only have rows that have count columns
allMeta_dt <- allMeta_dt[ID %in% colnames(counts_dt),]

# Sort rows to match count columns
allMeta_dt <- allMeta_dt[match(setdiff(colnames(counts_dt), geneCol_v), ID),]

# Double check
if (!identical(allMeta_dt$ID, setdiff(colnames(counts_dt), "gene"))) stop("Something is wrong with metadata rows and/or count columns.\n")
```

DESeq also doesn't like dashes and spaces in variables, so clean those as well.

```{r}
# Change Post Run-in to Post_RunIn and Week 8 to Week_8
allMeta_dt$Timepoint <- gsub(" ", "_", gsub("\\-in", "In", allMeta_dt$Timepoint))

# Change Non-responder to NonResponder
allMeta_dt$RECIST <- gsub("\\-r", "R", allMeta_dt$RECIST)
```

# Output

Write out cleaned data

```{r}
write.csv(counts_dt, file = file.path(dataDir_v, "cleanCounts.csv"), row.names = F, quote = F)
write.csv(allMeta_dt, file = file.path(dataDir_v, "cleanMetadata.csv"), row.names = F, quote = F)
```

